Beyond clines: lineages and haplotype blocks in hybrid zones

### 2016-04-26 drafts/hybrid_zone.tex
reworked the intro

\abstract{
Hybrid zones formed between recently diverged populations offer an opportunity 
to study the mechanisms underlying reproductive isolation and the process of speciation.
Here, we use a combination of analytical theory and explicit forward simulations
to describe how selection against hybrid genotypes impacts patterns of introgression
across genomic and geographic space. 
By understanding how lineages move across the hybrid zone, in a model without coalescence, 
we describe how clines form and how parental haplotypes are broken up during introgression.
Clines form in about $1/s$ generations, where $s$ is the strength of selection against hybrids,
and linked clines persist over a  genomic scale of $1/T$, where $T$ is the age, in generations, of the hybrid zone. 
Locally disadvantageous alleles tend to exist as small families,
whose lineages trace back to the side from which they originated at speed $\sqrt{s}$ dispersal distances per generation.
The lengths of continuous tracts of ancestry provide an additional source of information:
blocks of ancestry surrounding single-locus incompatibilities can be substantially longer than the genome-wide average block length at the same spatial location,
an observation that might be used to characterize the age of hybrid zones
and identify candidate targets of selection.
%\alisa{ascertainment bias. Tract lengths provide additional source of information, as they provide unbiased information even when used ascertained markers.}

@@

The process of speciation commonly involves populations diverging with relatively little gene flow \citep{Coyne2004}.  
% Gene flow between formerly isolated populations is possible if reproductive isolation is not complete.
However, when formerly isolated populations come into contact before reproductive isolation is complete, some gene flow is possible.
%As two such populations interbreed , each portion of each individual's genome 
%can be labeled according to which ancestral population is is inherited from, and local 
Interbreeding and migration between such populations  
% YB It seems like the fact that such ancestry can be labeled is more important later, but probably shouldn't be a focus here. 
creates a gradient of alleles derived from the two populations across geographic space, %YB removed `assignments' as it seems like it is the ancestry, not its assignment, that is of interest
centered on their point of contact \citep[reviewed by][]{Barton1985}. 
If the populations are sufficiently diverged, 
this process leaves a distinct pattern of variation across the genome, 
in which long tracts of divergent haplotypes from each ancestral population  
are broken up by historical recombination events,
forming chromosomal {junctions} in ancestry \citep{Fisher1954, Chapman2002, baird2003distribution}. 
% Limited opportunities for recombination in recent hybrid zones generates substantial correlations in ancestry (and hence, genotype) between distant markers, and
These patterns of correlated ancestry in admixed populations have been previously used to infer histories of hybridization and admixture \citep[e.g.][]{Gravel2012,Hellenthal2014,sedghifar2015spatial}.

The mixing of genomes across space, combined with selection, results in the formation of genetic clines. 
These patterns have been the traditional focus of both empirical and theoretical studies into hybrid zones. 
While here we focus on hybrid zones maintained by selection against hybrids 
-- \emph{tension zones}, as reviewed in \citet{Barton1985} --  % removed \citet{Bazykin1969}, it's cited later
clines may also be due to differential selection across environmental gradients \citep{Haldane1948,fisher1950frequencies,hanson1966effects},
% generating \emph{ecotones}.  % I think ecotone means something somewhat different: see Barton & Hewitt 1985, p. 132, for instance
Although these are in principle different,  \linelabel{ll:all_clines_are_the_same}
theoretical and simulation studies show that
genomic clines seen in the two situations are effectively indistinguishable without additional information \citep{may1975gene,Barton1993,Kruuk1999},
with their shape determined mostly by the mean fitness of the heterozygote \citep{Slatkin1973}.
% consistently show ecotones and tension zones have very similar properties \citep{Slatkin1973, barton1986barrier,Kruuk1999}, 
% \plr{(I can't find discussion of similarity of different types of clines in \citet{barton1986barrier}: put the citation back in if I missed it.)}
%% don't think we need this next bit
% The primary difference being that the center of a tension zone fluctuates over time \citep{barton1979dynamics} 
% because stochastic changes in allele frequency can shit the cline center, 
% while centers of ecotones do not change unless the environmental gradient changes.

Our work here builds on the large body of theory describing tension zones. 
While much of this previous work describes equilibrium states of genetic clines, 
we focus on \emph{transients}, in particular describing haplotype patterns. 
Although the cline in selected allele frequency establishes quickly, transient patterns can persist for quite some time, 
and haplotypes can potentially carry substantial information about the formation of recent zones.
To do this, we take the reverse-time perspective,
understanding the temporal dynamics of the system by describing properties of the lineages tracing back from the zone.
% This is particularly suited to understanding transient patterns.

The arrangement of segments of genome inherited from particular ancestral populations,
which may be identifiable as distinct haplotypes,
is shaped by the number of recombination events that have taken place between ancestral genotypes. 
Therefore the length distribution of such segments contains clues concerning the age of the hybrid zone and the relative strength of selection affecting subsets of the genome. 
The distribution of ancestry tract length in admixed or hybridizing populations has been used to infer the extent of hybridization and timing of secondary contact both in single populations and in hybrid zones \citep{Price2009, Gravel2012,sedghifar2015spatial}. In addition to allowing researchers to interpret non-equilibrium patterns of diversity in hybrid zones, 
the length distribution of ancestry tracts could reveal the impact of ongoing selection on the mixing of ancestral genomes. %and, consequently, the maintenance of species distinctness. 
%In the extreme case of zero fitness in first-generation hybrids, for example, parental genomes will never mix and the stable hybrid zone will resemble a recently formed one. 
%If hybrids have non-zero fitness, some amount of introgression can occur, and the distribution of ancestry tract length in such populations has been used to infer the extent %of hybridization and timing of secondary contact \citep{Price2009, Gravel2012}.

@@

Using a combination of theory and simulated hybrid zones, we describe genome-wide patterns of coancestry 
as they relate to hybrid zone age, genetic distance from selected locus and geographic distance from hybrid zone center. 
In \citet{sedghifar2015spatial} we described how lineages move across a neutral zone of secondary contact,
obtaining predictions for distributions of lengths of blocks of ancestry, which we fit to several datasets.
In this work, we extend the same framework to the case of selection against heterozygotes at a single locus,
by describing how lineages move conditional on the frequencies of the selected allele
and allowing lineages to recombine between selected background 
(analogous to the structured coalescent applied to selection, as in \citet{hudson1988coalescent}).

@@

\yb{Because of the well-known problems with spatial coalescence \citep{felsenstein1975pain,barton2002continuous},
incorporating genetic drift is not straightforward, and we therefore neglect genetic drift in our analytical results.} \revpoint{2}{9}
This approach is common to nearly all theoretical work,
that study the reaction-diffusion equations that govern the deterministic, high-density limit \citep[as in][]{Nagylaki1975},
or the discrete analogues \citep{hanson1966effects}. 
There is substantial work on how clines are affected by genetic drift \citep{Slatkin1975,felsenstein1975genetic,durrett2007width,barton2008effect,Polechova2011},
and quantifying the barrier to gene flow caused by clines \citep{barton1979gene,Barton1989,Barton1993,Barton2000}.
\plr{(why are references to 'barrier to gene flow' in the paragraph on drift? perhaps move to discussion?)}
While we do not attempt to account for finite population size in our analytic results,
our simulation results (below) suggest that although genetic drift strongly affects local noisiness of the system,
it does not substantially influence our conclusions at moderate population densities
(hundreds within one dispersal distance),
perhaps because of the short time-scales that we examine.
\plr{(maybe remove this next sentence? but want to say something about the very low densities we don't examine.)}
At much lower population densities, local extinction and recolonization are expected to have a stronger effect.

\plr{(nice wrap-up sentence shouldn't be at the end of paragraph on drift:)}
\alisa{Our results present a possible approach for identifying genomic regions experiencing selection in hybrid zones.}

@@

%%%%%%%%
\begin{figure}
\includegraphics[clip,width=\textwidth,trim=0cm 8cm 0cm 0.05cm]{figs/BM_schematic_new2.pdf}
\caption{
    \textbf{A schematic figure of the lineages along which a segment of genome has been inherited,}
    showing the location (horizontal axis) and time (vertical axis) at which the ancestors 
    of a hypothetical sampled haplotype lived.
    At the time of sampling ($\tau=0$), the segment contains the $B$ allele at the selected locus.
    Paths traced by lineages across space are depicted by blue and yellow lines. 
    Branching events occur when there was a recombination event within the sampled segment.
    The thick blue line represents the path of the selected locus. 
    \plr{(All lines look the same width: no thick and thin.)}
    The colors of the other paths indicate the identity at the linked selected locus 
    (blue: linked to allele $B$; yellow: linked to allele $A$). 
    The center of the hybrid zone is at geographic position $x=0$; in the region $x<0$, ancestry $B$ is less common,
    and because of selection, the selected lineage spends little time there.
  %  Red dots represent ancestral recombination events that fall in the given segment of genome in an ancestor;
  %  these can cause the lineage to branch (lineages at linked sites are depicted as colored lines),
  %  or can switch lineages at linked sites between selected backgrounds 
  %  (yellow lineages are linked to the $A$ selected locus; blue are linked to $B$).
  %  Here the initial haplotype is broken up into 3 chunks over 100 generations,
  %  which switch between selected backgrounds once.
    Based on position of the lineages at $\tau=100$ generations ago,
    from left to right the segments are of ancestry $(A,B,B)$.
    % \plr{(Add blue/yellow bars at the bottom?)}
}\label{Fig:schematic}
\end{figure}

>>>>

\abstract{
Hybrid zones formed between recently diverged populations offer an opportunity 
to study the mechanisms underlying reproductive isolation and the process of speciation.
Here, we use a combination of analytical theory and explicit forward simulations
to describe how selection against hybrid genotypes impacts patterns of introgression
across genomic and geographic space. 
By understanding how lineages move across the hybrid zone, in a model without coalescence, 
we describe how clines form and how parental haplotypes are broken up during introgression.
Clines form in about $1/s$ generations, where $s$ is the strength of selection against hybrids,
and linked clines persist over a  genomic scale of $1/T$, where $T$ is the age, in generations, of the hybrid zone. \alisa{This is in May1975? Should it be removed?}
Locally disadvantageous alleles tend to exist as small families,
whose lineages trace back to the side from which they originated at speed $\sqrt{s}$ dispersal distances per generation.
The lengths of continuous tracts of ancestry provide an additional source of information:
blocks of ancestry surrounding single-locus incompatibilities can be substantially longer than the genome-wide average block length at the same spatial location,
an observation that might be used to characterize the age of hybrid zones
and identify candidate targets of selection.
%\alisa{ascertainment bias. Tract lengths provide additional source of information, as they provide unbiased information even when used ascertained markers.}

@@

The process of speciation commonly involves populations diverging with relatively little gene flow \citep{Coyne2004}.  
% Gene flow between formerly isolated populations is possible if reproductive isolation is not complete.
However, when formerly isolated populations come into contact before reproductive isolation is complete, some gene flow is possible.
%As two such populations interbreed , each portion of each individual's genome 
%can be labeled according to which ancestral population is is inherited from, and local 
Interbreeding and migration between such populations  
% YB It seems like the fact that such ancestry can be labeled is more important later, but probably shouldn't be a focus here. 
creates a gradient of alleles derived from the two populations across geographic space, %YB removed `assignments' as it seems like it is the ancestry, not its assignment, that is of interest
centered on their point of contact \citep[reviewed by][]{Barton1985}. 
If the populations are sufficiently diverged, 
this process leaves a distinct pattern of variation across the genome, 
in which long tracts of divergent haplotypes from each ancestral population  
are broken up by historical recombination events,
forming chromosomal {junctions} in ancestry \citep{Fisher1954, Chapman2002, baird2003distribution}. 
% Limited opportunities for recombination in recent hybrid zones generates substantial correlations in ancestry (and hence, genotype) between distant markers, and
These patterns of correlated ancestry in admixed populations have been previously used to infer histories of hybridization and admixture \citep[e.g.][]{Gravel2012,Hellenthal2014,Harris2014,sedghifar2015spatial}.

%The mixing of genomes across space, combined with selection, results in the formation of genetic clines. 
%These patterns have been the traditional focus of both empirical and theoretical studies into hybrid zones. 
%While here we focus on hybrid zones maintained by selection against hybrids 
-%- \emph{tension zones}, as reviewed in \citet{Barton1985} --  % removed \citet{Bazykin1969}, it's cited later
%clines may also be due to differential selection across environmental gradients \citep{Haldane1948,fisher1950frequencies,hanson1966effects},
%% generating \emph{ecotones}.  % I think ecotone means something somewhat different: see Barton & Hewitt 1985, p. 132, for instance
%Although these are in principle different,  \linelabel{ll:all_clines_are_the_same}
%theoretical and simulation studies show that
%genomic clines seen in the two situations are effectively indistinguishable without additional information \citep{may1975gene,Barton1993,Kruuk1999},
%with their shape determined mostly by the mean fitness of the heterozygote \citep{Slatkin1973}.
% consistently show ecotones and tension zones have very similar properties \citep{Slatkin1973, barton1986barrier,Kruuk1999}, 
% \plr{(I can't find discussion of similarity of different types of clines in \citet{barton1986barrier}: put the citation back in if I missed it.)}
%% don't think we need this next bit
% The primary difference being that the center of a tension zone fluctuates over time \citep{barton1979dynamics} 
% because stochastic changes in allele frequency can shit the cline center, 
% while centers of ecotones do not change unless the environmental gradient changes.

Our work here builds on the large body of theory describing tension zones \alisa{and aims to describe \emph{transient} dynamics with a focus on \emph{haplotype} patterns.}
%While much of this previous work describes equilibrium states of genetic clines, 
%we focus on \emph{transients}, in particular describing haplotype patterns. 
%Although the cline in selected allele frequency establishes quickly, transient patterns can persist for quite some time, 
%and haplotypes can potentially carry substantial information about the formation of recent zones.
To do this, we take the reverse-time perspective,
understanding the temporal dynamics of the system by describing properties of the lineages tracing back from the zone.
% This is particularly suited to understanding transient patterns.

The arrangement of segments of genome inherited from particular ancestral populations%, which may be identifiable as distinct haplotypes,
is shaped by the number of recombination events that have taken place between ancestral genotypes. 
The length distribution of such segments therefore contains information about the age of the hybrid zone and the relative strength of selection affecting subsets of the genome. 
%The distribution of ancestry tract length in admixed or hybridizing populations has been used to infer the extent of hybridization and timing of secondary contact both in single populations and in hybrid zones \citep{Price2009, Gravel2012,sedghifar2015spatial}. In addition to allowing researchers to interpret non-equilibrium patterns of diversity in hybrid zones, 
%the length distribution of ancestry tracts could reveal the impact of ongoing selection on the mixing of ancestral genomes. %and, consequently, the maintenance of species distinctness. 
%In the extreme case of zero fitness in first-generation hybrids, for example, parental genomes will never mix and the stable hybrid zone will resemble a recently formed one. 
%If hybrids have non-zero fitness, some amount of introgression can occur, and the distribution of ancestry tract length in such populations has been used to infer the extent %of hybridization and timing of secondary contact \citep{Price2009, Gravel2012}.

@@

Using a combination of theory and simulated hybrid zones, we describe genome-wide patterns of coancestry 
as they relate to hybrid zone age, genetic distance from selected locus and geographic distance from hybrid zone center. 
\alisa{We extend the framework presented by \cite{Sedghifar2015} to describe how lineages move across a zone of secondary contact. We consider the influence of selection against heterozygotes at a single locus on the movement of lineages across space, conditioning on the frequencies of the selected allele and allowing lineages to recombine between selected background } (analogous to the structured coalescent applied to selection, as in \citet{hudson1988coalescent}).
%In \citet{sedghifar2015spatial} we described how lineages move across a neutral zone of secondary contact,
%obtaining predictions for distributions of lengths of blocks of ancestry, which we fit to several datasets.
%In this work, we extend the same framework to the case of selection against heterozygotes at a single locus,
%by describing how lineages move conditional on the frequencies of the selected allele
%and allowing lineages to recombine between selected background 
%(analogous to the structured coalescent applied to selection, as in \citet{hudson1988coalescent}).

@@

\yb{Because of the well-known problems with spatial coalescence \citep{felsenstein1975pain,barton2002continuous},
incorporating genetic drift is not straightforward, and we therefore neglect genetic drift in our analytical results.} \revpoint{2}{9}
This approach is common to nearly all theoretical work,
that study the reaction-diffusion equations that govern the deterministic, high-density limit \citep[as in][]{Nagylaki1975},
or the discrete analogues \citep{hanson1966effects}. 
There is substantial work on how clines are affected by genetic drift \citep{Slatkin1975,felsenstein1975genetic,Nagylaki1978,durrett2007width,barton2008effect,Polechova2011},
and quantifying the barrier to gene flow caused by clines \citep{barton1979gene,Barton1989,Barton1993,Barton2000}.
\plr{(why are references to 'barrier to gene flow' in the paragraph on drift? perhaps move to discussion?)}
While we do not attempt to account for finite population size in our analytic results,
our simulation results (below) suggest that although genetic drift strongly affects local noisiness of the system,
it does not substantially influence our conclusions at moderate population densities
(hundreds within one dispersal distance),
perhaps because of the short time-scales that we examine.
\plr{(maybe remove this next sentence? but want to say something about the very low densities we don't examine.)}
At much lower population densities, local extinction and recolonization are expected to have a stronger effect.

\plr{(nice wrap-up sentence shouldn't be at the end of paragraph on drift:)} \alisa{Is it weird to end intro on above paragraph? Could this wrap-up sentence hold up on its own? }
\alisa{Our results look beyond the traditional signature of selection observed in the width of genetic clines, and present a possible new approach for identifying genomic regions experiencing selection in hybrid zones.}

@@

%%%%%%%%
\begin{figure}
\includegraphics[clip,width=\textwidth,trim=0cm 8cm 0cm 0.05cm]{figs/BM_schematic_new2.pdf}
\caption{
    \textbf{A schematic figure of the lineages along which a segment of genome has been inherited,}
    showing the location (horizontal axis) and time (vertical axis) at which the ancestors 
    of a hypothetical sampled haplotype lived.
    At the time of sampling ($\tau=0$), the segment contains the $B$ allele at the selected locus.
    Paths traced by lineages across space are depicted by blue and yellow lines. 
    Branching events occur when there was a recombination event within the sampled segment.
    The \alisa{rightmost} blue line represents the path of the selected locus. 
   % \plr{(All lines look the same width: no thick and thin.)}
    The colors of the other paths indicate the identity at the linked selected locus 
    (blue: linked to allele $B$; yellow: linked to allele $A$). 
    The center of the hybrid zone is at geographic position $x=0$; in the region $x<0$, ancestry $B$ is less common,
    and because of selection, the selected lineage spends little time there.
  %  Red dots represent ancestral recombination events that fall in the given segment of genome in an ancestor;
  %  these can cause the lineage to branch (lineages at linked sites are depicted as colored lines),
  %  or can switch lineages at linked sites between selected backgrounds 
  %  (yellow lineages are linked to the $A$ selected locus; blue are linked to $B$).
  %  Here the initial haplotype is broken up into 3 chunks over 100 generations,
  %  which switch between selected backgrounds once.
    Based on position of the lineages at $\tau=100$ generations ago,
    from left to right the segments are of ancestry $(A,B,B)$.
    % \plr{(Add blue/yellow bars at the bottom?)}
}\label{Fig:schematic}
\end{figure}
